---
title: "Monte Carlo approximation of π"
description: |
  A short description of the post.
author:
  - name: Yevgen Ryeznik
date: 02-16-2019
output:
  distill::distill_article:
    self_contained: false
    toc: TRUE
    toc_float: TRUE
categories:
  - R
  - Scientific computing
  - Monte Carlo
  - Statistical simulations  
---


```{r setup, include=FALSE}

# setting chunk options
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
knitr::opts_chunk$set(fig.align = "center", fig.width = 6, fig.height = 4)

# loading libraries
library(tidyverse)

# auxiliary functions
add_theme <- function(){
  theme_bw()+
  theme(
    axis.text = element_text(family="Helvetica", face = "bold", size = 10),
    axis.title = element_text(family="Helvetica", face = "bold", size = 12),
    strip.text = element_text(family="Helvetica", face = "bold", size = 12),
    title = element_text(family="Helvetica", face = "bold", size = 14),
    #legend.position = "right",
    legend.text = element_text(family="Helvetica", face = "bold", size = 12)
  )
}
```

## Approximation of $\pi$ by geometric probabilities

Let us consider a quarter of a circle inscribed in a square with a side length equal to 1 (Fig. \@ref(fig:geometry))

```{r geometry, echo=FALSE, fig.cap="Geometry of the problem."}

square <- tibble(
  x = c(0, 0, 1, 1, 0),
  y = c(0, 1, 1, 0, 0)
)

circle <- tibble(
  t = seq(0, 0.5, length = 100)*pi,
  x = cos(t),
  y = sin(t)
)

ggplot()+
  geom_path(data = square, aes(x, y))+
  geom_line(data = circle, aes(x, y))+
  add_theme()
```

A point is dropped inside the square. The question is: _what is the probability that a point will be **inside** the circle?_

The answer to this question is given by the formula of geometric probabilities:

$$
  \text{Pr}\left(\text{point belongs to the circle}\right) = \frac{\text{area of the quarter of the circle}}{\text{area of the square}} = \frac{\frac{\pi}{4}}{1} = \frac{\pi}{4}.
$$
At the same time, if we drop a large number ($n$) of points inside the square, there will be some number of points ($m$ ($m \leq n$) inside the quarter of the circle. And the probability above can be calculated by the classical probability formula: 

$$
  \text{Pr}\left(\text{point belongs to the circle}\right) = \frac{\pi}{4} \approx \frac{\text{number of points inside the quarter of the circle}}{\text{total number of points}}  = \frac{m}{n},
$$
Thus, we obtain an approximation for $\pi$ as

$$
  \pi \approx 4\frac{m}{n}
$$


## Implementation in R

The following code performs the approximation. The code is taken from [Bruno Rodrigues' blog](https://www.brodrigues.co/blog/2018-12-21-tidyverse_pi/)

```{r approximation, message=FALSE}
set.seed(12345)

# number of simulations
n <- 100000

# approximation table
approx <- tibble(
  x = runif(n),
  y = runif(n),
  inside = as.integer(x^2 + y^2 < 1)
) %>%
  rowid_to_column("n") %>% 
  mutate(pi = 4*cumsum(inside)/n)
```

Let us take a look at the last 20 rows in the approximation table (Tab. \@ref(tab:approx-tab)):

```{r approx-tab, echo=FALSE}
approx[(n-19):n, ] %>% 
  knitr::kable(caption = "Last 20 approximate values of $\\pi$ in the simulation.")
```

Now, let us visualize the experiment for $n = 100, 500, 1000, 10000$ (Fig. \@ref(fig:visualization)).

```{r visualization, fig.cap="Visualization of experiments for $n \\in \\{100, 500, 1000, 10000\\}$.", preview=TRUE}
# visualization
square <- tibble(
  x = c(0, 0, 1, 1, 0),
  y = c(0, 1, 1, 0, 0)
)

circle <- tibble(
  t = seq(0, 0.5, length = 100)*pi,
  x = cos(t),
  y = sin(t)
)

n_ <- c(100, 500, 1000, 10000)
label <- map_chr(n_, ~ str_glue("n == {.x}: pi %~~% {approx_val}", approx_val = approx$pi[.x]))

map_df(n_, ~ {
  filter(approx, n <= .x) %>% 
    add_column(title = .x)
}) %>% 
  mutate(title = factor(title, labels = label)) %>% 
  ggplot() +
    geom_point(aes(x, y, color = factor(inside)))+
    geom_path(data = square, aes(x, y))+
    geom_line(data = circle, aes(x, y))+
    scale_color_manual(values = c("darkblue", "red"))+
    facet_wrap(~ title, labeller = label_parsed)+
    theme_bw()+
    theme(
      axis.text = element_text(family="Helvetica", face = "bold", size = 10),
      axis.title = element_text(family="Helvetica", face = "bold", size = 12),
      strip.text = element_text(family="Helvetica", face = "bold", size = 12),
      title = element_text(family="Helvetica", face = "bold", size = 14),
      legend.position = "none"
  )
```

Finally, let us take look at the convergence plot (Fig. \@ref(fig:convergence)).

```{r convergence, fig.cap = "Convergence of the Monte Carlo Approach"}
ggplot(approx, aes(n, pi))+
  geom_line(aes(n, pi), color = "red")+
  geom_hline(yintercept = pi)+
  scale_x_log10()+
  labs(
    x = "number of samples",
    y = "approximate value"
  )+
  theme_bw()+
  theme(
    axis.text = element_text(family="Helvetica", face = "bold", size = 10),
    axis.title = element_text(family="Helvetica", face = "bold", size = 12),
    strip.text = element_text(family="Helvetica", face = "bold", size = 12),
    title = element_text(family="Helvetica", face = "bold", size = 14)
  )
```



