---
title: "Pythagorean theorem day"
description: |
  What is a "Pythagorean theorem day"? Calculation Pythagorean theorem days in the XXI century.
author:
  - name: Yevgen Ryeznik
date: 02-29-2020
output:
  distill::distill_article:
    self_contained: false
    toc: TRUE
    toc_float: TRUE
categories:
  - Python
  - Pythagorean theorem
---


```{r setup, include=FALSE}

# setting chunk options
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
knitr::opts_chunk$set(fig.align = "center", fig.width = 7, fig.height = 5)

library(tidyverse)
```

## Pythagorean's theorem

In mathematics, the _**Pythagorean theorem**_, gives relation in Euclidean geometry among the three sides of a _**right triangle**_. It states that the area of the square whose side is the hypotenuse (the side opposite the right angle) is equal to the sum of the areas of the squares on the other two sides. 

It can be written in the form of the following equation:

$$
  a^2+b^2 = c^2.
$$

```{r right-triangle-geometry, echo=FALSE}
triangle <- tibble(
  x = c(0, 4, 0, 0),
  y = c(0, 0, 3, 0)
)

square1 <- tibble(
  x = c(-3, 0, 0, -3, -3),
  y = c( 0, 0, 3,  3,  0),
  ymin = rep(0, length(y)),
  ymax = rep(3, length(y))
) 

square2 <- tibble(
  x = c( 0, 4,  4,  0,  0),
  y = c( 0, 0, -4, -4,  0),
  ymin = rep(-4, length(y)),
  ymax = rep(0, length(y))
) 

square3 <- tibble(
  x = c( 0, 3,  7,  4,  0),
  y = c( 3, 7,  4,  0,  3),
)

square3_fill <- tibble(
     x = c(0, 3, 4, 7),
  ymin = c(3, 0.85, 0, 4),
  ymax = c(3, 7, 6.25, 4)
)

tbl_text <- tibble(
  x = c(-1.5,  2, 3.5, 0.2, 1.5, 1.8),
  y = c( 1.5, -2, 3.5, 1.5, 0.2, 1.3),
  txt = c("a^2", "b^2", "c^2", "a", "b", "c")
)

ggplot()+
  geom_path(data = triangle, aes(x, y))+
  geom_path(data = square1, aes(x, y))+
  geom_ribbon(data = square1, aes(x = x, ymin = ymin, ymax = ymax), fill = "red", alpha = 0.5)+
  geom_path(data = square2, aes(x, y))+
  geom_ribbon(data = square2, aes(x = x, ymin = ymin, ymax = ymax), fill = "lightblue", alpha = 0.75)+
  geom_path(data = square3, aes(x, y))+
  geom_ribbon(data = square3_fill, aes(x = x, ymin = ymin, ymax = ymax), fill = "lightgreen", alpha = 0.75)+
  geom_text(data = tbl_text, aes(x, y, label = txt), parse = TRUE)+
  coord_fixed(ratio = 1)+
  labs(
    x = "", y = ""
  )+
  scale_x_continuous(breaks = c())+
  scale_y_continuous(breaks = c())+
  theme_bw()+
  theme(panel.border = element_blank())

```

## Pythagorean theorem day

I found a funny definition of what is a _**Pythagorean theorem day**_. It is the day which date in the format "yy-mm-dd" satisfies the following eqaulity:

$$
  \text{yy}^2 = \text{mm}^2 + \text{dd}^2.
$$

For example, August 15, 2017 can be written as 17-08-15, and it is a Pythagorean theorem day since 

$$
  17^2 = 8^2 + 15^2.
$$

I was curious how many days within the 21st century are the Pythagorean theorem days?

## Python code to find the Pythagorean theorem days

```{python finding-pythagorean-dates}
import datetime as dt

year = [2000 + y for y in range(100)]
month = range(1, 13)
day = range(1, 32)

dates = [(y, m, d) for y in year for m in month for d in day]

for date in dates:
  if (date[0]-2000)**2 == date[1]**2 + date[2]**2:
    try:
      x = dt.datetime(date[0], date[1], date[2])
      print(x.strftime("%Y-%m-%d"))
    except:
      pass
```

As we can see, only 3 such days left in the 21st century, and one will be this year on December 16, 2020.