---
title: "Lecture 6: Statistical inference -- Estimation. Part 2"
description: |
  Bayesian approach to the statistical inference is introduced.
author: Yevgen Ryeznik
institute: Department of Pharmaceutical Biosciences, Uppsala University.
date: 2019-11-29
output:
  distill::distill_article:
    self_contained: false
    toc: TRUE
    toc_float: TRUE
header-includes:
  - \usepackage{amsthm}
  - \usepackage{amsfonts}
  - \usepackage{hyperref}
  - \usepackage{url}
  - \usepackage{xcolor}
  - \usepackage{array}
  - \usepackage{booktabs}
  - \usepackage{colortbl}
  - \usepackage{float}
  - \usepackage{longtable}
  - \usepackage{makecell}
  - \usepackage{multirow}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage{wrapfig}
  - \usepackage[normalem]{ulem}
---


```{r,setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, message = FALSE, echo=FALSE)
knitr::opts_chunk$set(fig.align = "center", fig.width = 6, fig.height = 4)
options(knitr.table.format = "html")
options(kableExtra.latex.load_packages = FALSE)

library(tidyverse)
library(kableExtra)

add_theme <- function()
  theme_bw()+
  theme(
    axis.text = element_text(family="Helvetica", face = "bold", size = 10),
    axis.title = element_text(family="Helvetica", face = "bold", size = 12),
    strip.text = element_text(family="Helvetica", face = "bold", size = 12),
    title = element_text(family="Helvetica", face = "bold", size = 14),
    legend.position = "bottom",
    legend.text = element_text(family="Helvetica", face = "bold", size = 12)
  )
```

[**Download pdf**](inference-part2.pdf)

## Bayesian approach

Let $\theta$ be a parameter of interest. Somehow, a prior information about $\theta$ is available (historical reports, similar experiments, subjective estimates, etc. ). We can represent this information in terms of a prior distribution

$$
  p(\theta)
$$

Then, after performing an experiment where we observe a random variable $Y=y$. Bayes' formula tells that

$$
  p(\theta|y) = \frac{p(\theta,y)}{p(y)} = \frac{p(\theta)p(y|\theta)}{p(y)}                                   
$$

where $p(\theta,y)$ is a _**posterior distribution**_ and 

$$
\begin{aligned}
  p(y)&=\sum\limits_{\theta}p(\theta)p(y|\theta) \quad\text{(discrete case)} \\
  &\text{ or } \\
  p(y)&=\int\limits_{\Theta}p(\theta)p(y|\theta)d\theta \quad\text{ (continuous case)}
\end{aligned}
$$
Formulas can be rewritten as 

$$
  p(\theta|y)\propto p(\theta)p(y|\theta).
$$

_**Example 1**_ (Spelling correction)

What about the word “_radom_”? It could be:

- Misspelling or mistyping of “random”.
- Misspelling or mistyping of “radon”.
- Intentional typing of “radom”.

Label $y$ as the data (observation of $Y$) and $\theta$ as the word that a person was intending to type:

$$
  \text{Pr}(\theta|Y =\textit{"radom"})\propto p(\theta)\text{Pr}(Y=\textit{"radom"}|\theta).
$$
Consider only three possibilities for the intended word, $\theta$: (“random”, “radon”, “radom”). 

Posterior probability of interest, e.g. for $\theta_1$="random":

$$
\text{Pr}(\theta_1 |\textit{"radom"}) = \frac{p(\theta_1)\text{Pr}(\textit{"radom"}|\theta_1 )}{\sum_{j=1}^3p(\theta_j )\text{Pr}(\textit{"radom"}|\theta_j)}.
$$

We need:

- Prior probabilities $p(\theta_j)$ (frequencies of words in some large database)
- Likelihoods $p(y|\theta_j)$ (from model of spelling and typing errors)

**Remark**. _Radom_ is a city in Poland 

## Estimation of probability $p= \text{Pr}(B)$ of some event $B$

We want to study $p=p_B$. Two approaches can be applied:

- _**Frequentist**_.  $\widehat{p}_B=\frac{k}{n}$,  where $k$ is a number of times when $B$ is true in $n$ trials. The uncertainty on estimate can be done with confidence intervals.
- _**Bayesian**_. Uncertainty of $p_B$ is modeled by a p.d.f. $f(p)$. Then

$$
  \text{Pr}^\text{pred}(B)=\mathbf{E}[p]=\int pf(p)dp.
$$

_**The crucial point in Bayesian methodology is the choice of the density $f(\cdot)$. The density reflects all our knowledge about the studied problem**_.

## Combining data and prior knowledge.

- An experiment starts with a _**prior**_ assumption about the parameter of interest:

$$
  f^\text{prior}(\theta)
$$
- Then, a result of experiment is observed $\Leftrightarrow$ some event $C$ is true.
- Prior information is updated by the Bayes' formula:

$$
  f^\text{post}(\theta)=cP(C|\theta)f^\text{prior}(\theta)
$$

where

$$
  \frac{1}{c} = P(C)= \int_{-\infty}^{+\infty}P(C|\theta)f^\text{prior}(\theta)d\theta.
$$

_**Example 2**_ (severe adverse event (AE) in a clinical trial)

Let us consider a clinical trial where 1 severe AE out of 3 AEs has been observed, and we are interested in probability $p_B$ of a severe AE. Since we had no information about it before the experiment, we may assume that $p_B\sim U(0,1)$.

$$
  f^\text{prior}(p)=1.
$$

Now, we have observed a severe AE:

$$
  C = \text{"one severe AE out of 3 AEs"} \Leftrightarrow N = 1,\text{ }N\sim Bin(3, p_B)
$$

$$
\begin{aligned}
  P(N=1|p)&=P(C|p)= {3\choose 1} p(1-p)^2 \Rightarrow \\
  f^\text{post}(p) &= c\cdot 3p(1-p)^2\cdot 1=cp(1-p)^2, \text{ }0\leq p \leq 1,
\end{aligned}
$$

which is $Beta(2,3)$:

$$
\theta\sim Beta(a,b),\quad f(\theta) = c\theta^{a-1} (1-\theta)^{b-1},\quad a, b > 0, \theta\in (0,1).
$$

```{r beta-prior-posterior}
x <- seq(0, 1, by = 0.01)
prior <- rep(1, length(x))
posterior <- dbeta(x, 2, 3)

tibble(x, prior, posterior) %>% 
  gather(distribution, value, -x) %>% 
  ggplot(aes(x = x, y = value))+
  geom_line(aes(color = distribution), size = 1.5)+
  scale_color_manual(values = c("red", "darkblue"))+
  xlab(expression(paste("probability, ", italic(p))))+
  ylab("")+
  add_theme()
```

### Credibility Intervals

$(1-\alpha)\cdot 100\%$ credibility interval for the parameter $\theta$ is given by

$$
  \text{Pr}(\theta_{1-\alpha/2}\leq \theta \leq \theta_{\alpha/2} )=1-\alpha.
$$                                                                    

_**Example 3**_ (Normal Distribution)

$X_1,X_2,…,X_n \sim N(\mu, \sigma^2)$, and we are interested in estimating $\mu$:

- prior distribution $\mu\sim N(\mu_0,\sigma_0^2)$

It can be shown that the posterior for $\mu$ is also normal $N(\mu_1,\sigma_1^2)$ where

$$
 \mu_1=\frac{\frac{n}{\sigma^2}\overline{X}+\frac{1}{\sigma^2}\mu_0}{\frac{n}{\sigma^2} + \frac{1}{\sigma_0^2}}\qquad \sigma_1^2=\left(\frac{n}{\sigma^2}^2 + \frac{1}{\sigma_0^2}\right)^{-1}.
$$

95% credibility interval is

$$
  \mu_1-1.96\sigma_1 \leq \mu \leq \mu_1+1.96\sigma_1.
$$

## Choice of a Model for Data: Conditional Distribution

Important step in Bayesian modeling: assume that $F(x;\theta)$ is actually the conditional distribution of $x$ given $\theta$, i.e. 

$$
  \text{Pr}(X\leq x|\theta) = F(x;\theta)
$$

The p.d.f. or p.m.f. of the conditional distribution $F(x|\theta)$ is denoted by $f(x|\theta)$.

_**Example 4**_ (Tossing a coin)

Fair coin: $X\sim Bin(n,1/2)$.

Unknown coin: Distribution $P(X\leq x|\theta)$; the p.m.f. is $p_{X|\theta}(x|\theta)=Bin(n,\theta)$.

## Conditional Independence

The independence between variables is assumed to be valid for any value of $\theta$

$$
  \text{Pr}(X_1\leq x,\ldots, X_n\leq x) = F(x_1|\theta)\cdot\ldots\cdot F(x_n|\theta)
$$

for all $\theta$.

## Bayesian Updating and Likelihood Functions

Observations: $x_1,\ldots, x_n$ of $X_1,\ldots, X_n$.

Assume that $f(x|\theta)$ is known and that observations of $X_i$ are conditionally independent.

Then, 

$$
  f(x_1,\ldots, x_n|\theta) = f(x_1|\theta)\cdot\ldots\cdot f(x_n|\theta) = \mathcal{L}(\theta)
$$

is a likelihood.

$$
  f^\text{post}(\theta)=c\mathcal{L}(\theta)f^\text{prior}(\theta), \qquad  \frac{1}{c}= \int_{-\infty}^{+\infty}\mathcal{L}(\theta) f^\text{prior}(\theta)d\theta.
$$

## Conjugate Priors

A conjugate prior is an algebraic convenience, giving a closed-form expression for the posterior; otherwise a difficult numerical integration may be necessary. 

Further, conjugate priors may give intuition, by more transparently showing how a likelihood function updates a distribution. 

### Beta distribution

$$
\begin{array}{c}
\theta\sim Beta(a,b),a,b>0, \text{ and }f(\theta) =c\theta^{a-1} (1-\theta)^{b-1},\text{ }0\leq\theta\leq 1, c=\frac{\Gamma(a+b)}{\Gamma(a)\Gamma(b)} \\   
E[\theta]=p,V[\theta]=\frac{p(1-p)}{(a+b+1)},\text{ where } p=\frac{a}{a+b}.
\end{array}
$$

The Beta priors are conjugated priors for the problem of estimating the probability $p = \text{Pr}(B)$. 

- Observed in $n$ tries: the statement $B$ was true $k$ times. 
- Prior density: $f^\text{prior}(\theta) \sim Beta(a,b)$ 
- Posterior density follows as with 

$$
  f^\text{post} = Beta(\widetilde{a}, \widetilde{b}), \text{ where } \widetilde{a} = a + k, \widetilde{b} = b + n-k.
$$

The $Beta(a,b)$ prior is equivalent to observing that the event $B$ has occurred $a$ times in $a + b$ experiments. 


### Dirichlet distribution

$$
\begin{array}{c}
  \theta=(\theta_1, \theta_2)\sim Dirichlet(a_1,a_2,a_3), a_i > 0 \\
  f(\theta_1, \theta_2) = c\theta_1^{a_1-1}\theta_2^{a_2-1}(1-\theta_1-\theta_2)^(a_3-1), \theta_i>0, \theta_1 + \theta_2 < 1 \\ 
c = \frac{\Gamma(a_1+a_2+a_3)}{\Gamma(a_1)\Gamma(a_2)\Gamma(a_3)} \\ 
\mathbf{E}[\theta_i] = \frac{a_i}{a_0} ,\mathbf{V}[\theta_i]=\frac{a_i(a_0-a_i )}{a_0^2(a_0+1)}, \text{ where } a_0=a_1+a_2+a_3, i=1,2.
\end{array}
$$

The marginal distributions: $\theta_i\sim Beta(a_i,a_0-a_i)$, $i=1,2$.

The Dirichlet priors are conjugated priors for the problem of estimating the probabilities 
$p_i = \text{Pr}(B_i)$, $i = 1,2,3$, such that $B_i$ are disjoint and $p_1  + p_2  + p_3  = 1$. 

Let $\theta_i= p_i$.

- Observed in $n$ tries: the statement $B_i$  was true $k_i$ times. 
-	The prior density $f^\text{prior}(\theta_1, \theta_2) \sim Dir(a_1,a_2,a_3)$. 
- Posterior density: 

$$
f^\text{post}(\theta_1,\theta_2) \sim Dir(\widetilde{a}_1, \widetilde{a}_2, \widetilde{a}_3), \text{ where } \widetilde{a}_1 = a_1 + k_1, \widetilde{a}_2 = a_2 + k_2, \widetilde{a}_3 = a_3 + k_3. 
$$

### Gamma distribution

$$ 
\begin{array}{c}
  \theta\sim\Gamma(a,b), a, b> 0, \\
  f(\theta) = c\theta^{a-1}(1-\theta)^{b-1}, 0\leq\theta\leq 1, \\
  c=\frac{\Gamma(a+b)}{\Gamma(a)\Gamma(b)} \\
  \mathbf{E}[\theta]=\frac{a}{b}, \mathbf{V}[\theta]=\frac{a}{b^2}.
\end{array}
$$

The Gamma priors are conjugated priors for the problem of estimating the Poisson $\lambda$. 

- Observed: Observed in $n$ tries (Poisson experiment): the statement $B$ was true $k$ times.
- Prior density: $f^\text{prior}(\theta)\sim \Gamma(a,b)$.
- Posterior density:

$$ f^\text{post}(\theta) \sim \Gamma(\widetilde{a}, \widetilde{b}), \text{ where } \widetilde{a} = a + k, \widetilde{b} = b + n.                        
$$    


