---
title: "Galerkin method of solving the 1st kind hypersingular integral equation for 2D-scattering problem."
description: |
  Here, we present the Galerkin method of moments (GMoM) to solve an integral equation of the 1st kind with a hypersingular kernel. 
author:
  - name: Yevgen Ryeznik
date: 11-05-2015
categories:
  - Galerkin method
  - Integral equations
  - Mathematical physiscs
  - Python
output:
  distill::distill_article:
    self_contained: false
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", fig.width = 6, fig.height = 4)

library(ggplot2)
library(magrittr)
```

## Problem formulation

Let us consider a scattering problem of a plane monochromatic wave from a 2D screen (Fig. \@ref(fig:geometry)).

```{r geometry, fig.cap="Geometry of the problem."}
t <- seq(0, 2, length.out = 100)*pi
alpha <- 40*pi/180
A <- 1-alpha/(2*pi)
B <- alpha/2

x <- cos(A*t + B)
y <- sin(A*t + B)

tibble::tibble(x, y) %>% 
  ggplot()+
    geom_path(aes(x, y))+
    geom_curve(
      aes(x = 0, y = -2, xend = 0, yend = 2),
      arrow = arrow(length = unit(0.05, "npc")),
      curvature = 0
    )+
    geom_curve(
      aes(x = -2, y = 0, xend = 2, yend = 0),
      arrow = arrow(length = unit(0.05, "npc")),
      curvature = 0
    )+
    geom_curve(
      aes(x = 2*cos(5*pi/4), y = 2*sin(5*pi/4), xend = 1.1*cos(5*pi/4), yend = 1.1*sin(5*pi/4)),
      arrow = arrow(length = unit(0.05, "npc")),
      curvature = 0
    )+
    geom_curve(
      aes(x = 1.1*cos(5*pi/4), y = 1.1*sin(5*pi/4), xend = 0, yend = 0),
      curvature = 0, 
      linetype = "dashed"
    )+
    geom_curve(
      aes(x = 0.25*cos(5*pi/4), y = 0.25*sin(5*pi/4), xend = 0, yend = -0.25),
      curvature = 0.5
    )+
    geom_text(aes(2, -0.1), label = "x", size = 6)+
    geom_text(aes(0.1,  2), label = "y", size = 6)+
    geom_text(aes(1.1*cos(7*pi/4), 1.1*sin(7*pi/4)), label = "Gamma", size = 6, parse = TRUE)+
    geom_text(aes(1.5*cos(5*pi/4), 1.5*sin(5*pi/4)), label = "italic(bold(u))[0]", size = 6, parse = TRUE, vjust = -0.5)+
    geom_text(aes(0.25*cos(11*pi/8), 0.25*sin(11*pi/8)), label = "italic(i)[0]", size = 6, parse = TRUE, vjust = 1.5)+
    theme_bw()+
    theme(
      axis.text = element_blank(),
      axis.title = element_blank(),
      axis.ticks = element_blank(),
      panel.border = element_blank()
    )

```

Here,

-   $\Gamma$ is the screen boundary.
-   $\boldsymbol{u}_0$ is an initial incidence field.
-   $i_0$ is an incidence angle.

The question is: *what is the scattered field*?

## Integral equation-based approach to solve the problem

Let us consider a case of $H$-scattering (i.e. scattering of the magnetic field ).

Let $\boldsymbol{r} = \boldsymbol{r}(x, y)$ be a radius-vector of point on a coordinate plane. Then,

$$
H_0(\boldsymbol{r}) = e^{-jk\left(x\sin{(i_0)} + y\cos{(i_0)}\right)}
$$ is the incidence field, and

$$
H_{sc}(\boldsymbol{r}) = \int_{\Gamma}\frac{1}{4j}\frac{\partial}{\partial\boldsymbol{n}'}H_0^{(2)}\left(k\left|\boldsymbol{r}-\boldsymbol{r}'\right|\right)\psi(\boldsymbol{r'})dl'
$$

is the scattered field.

Here,

-   $j$ is the imaginary unit, i.e. $j^2 = -1$.
-   $k$ is a freespace wavenumber.
-   $H_0^{(2)}(z)$ is the [*Hankel function*](https://en.wikipedia.org/wiki/Bessel_function#Hankel_functions:_H(1)α,_H(2)α) of the $2^\text{nd}$ kind and zero order.

According to the superposition principle, the total field is defined as

$$
H_{total}(\boldsymbol{r}) = H_{0}(\boldsymbol{r}) + H_{sc}(\boldsymbol{r}).
$$

Next step is
