---
title: "Calling Julia from R via JuliaCall"
description: |
  In this blog post, we demonstrate how to use a code written in Julia programming language in an R script by using R package JuliaCall.
author:
  - name: Yevgen Ryeznik
  - url: https://yevgen.github.io
date: 2020-11-20
categories:
  - Julia
  - R
output:
  distill::distill_article:
    self_contained: false
    toc: TRUE
    toc_float: TRUE
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE
)

```

## `R` package `JuliaCall` 

In [one of the previous blog posts](../2018-04-10-extending-r-with-julia/index.html), I wrote about the extending `R` with `Julia` by using [`XRJulia` `R` package](https://github.com/johnmchambers/XRJulia).

Here, I would like to talk about another possibility of using `Julia` code in `R` projects. For seamless integration of `R` and `Julia`, one can use an `R` package called [`JuliaCall`](https://non-contradiction.github.io/JuliaCall/index.html). 

To install the package run the command:

```{r, E, eval=FALSE}
install.packages("JuliaCall")
```

Once the package is installed, it is necessary to initialize `Julia` to use it in the current `R` session:

```{r}
library(JuliaCall)
julia_home <- paste0(Sys.getenv('JULIA_HOME'), '/bin')
julia_setup(JULIA_HOME = julia_home)
```

Let's explain the above code. To initialize `Julia` a function `julia_setup()` is used. It requires an input parameter `JULIA_HOME`, which contains a path to the `Julia` executable file. On my laptop, I have a `JULIA_HOME` system variable specified in the .Renviron file:

```
JULIA_HOME = '/Applications/Julia-1.5.app/Contents/Resources/julia/',
```

and `Julia` executable file is located in the folder `$JULIA_HOME/bin`. So, the command

```{r, eval=FALSE}
julia_home <- paste0(Sys.getenv('JULIA_HOME'), '/bin')
```

create a string containing the path to the `Julia` executable:

```
julia_home = '/Applications/Julia-1.5.app/Contents/Resources/julia/bin',
```
and this value is used as an input parameter value to the function `julia_setup()`.

If `Julia` is not installed on your computer, then download it from the [`Julia` web page](https://julialang.org/downloads/) or use the function `install_julia()` from the `JuliaCall` package:

```{r, eval = FALSE}
destination_folder <- # specify the installation path
install_julia(destination_folder)
```

If everything is OK, then we can start using `Julia` in our `R`project.


## Example: integral evaluation with Monte Carlo (MC) approach

Let us evaluate the following integral 

$$
  I = \int_0^{+\infty}{x^2e^{-\frac{x^2}{2}}}dx
$$

It can be rewritten as an expectation of the function 

$$
f(x)=\left\{
  \begin{array}{lr}
    \sqrt{2\pi}x^2, & x\geq 0 \\ 
    0, & x < 0
  \end{array}
  \right.
$$

with respect to the normal distribution $N(0, 1)$ with p.d.f. 

$$
  p(x)=\frac{1}{\sqrt{2\pi}}e^{-\frac{x^2}{2}} 
$$

Then,

$$
  I = \text{E}\left[f(x)\right] \approx \frac{1}{n}\sum_{i = 1}^n{f(x_i)},
$$

where $x_1, x_2, \ldots, x_n$ are the sample points from the standard normal distribution $N(0, 1)$.

Let us implement the latter formula by using both `R` and `R` extended with `Julia`.

1. `R` implementation:

```{r mc-R}

# First, let us create a function to integrate on the R side
fcnR <- function(x) {
  dplyr::if_else(x >= 0, sqrt(2*pi)*x^2, 0)
}

# Second, we implement a MC procedure on the R side 
integrate_mcR <- function(fcn, n, seed = 12345) {
  # sampling from N(0, 1)
  set.seed(seed)
  x <- rnorm(n)
  
  # mean of f(x)
  mean(fcn(x))
}

# Finally, we evaluate the integral (with n=100,000)
  IR <- integrate_mcR(fcnR, 100000)
```

2. `Julia` implementation:

We use a function `julia_command()` to run `Julia` commands on the `Julia` side. If we need a value to use in `R`, there is a function called `julia_eval()`.

```{r mc-Julia}
# First, let us create a function to integrate on the Julia side
  julia_command(
    "function fcn(x)
      if (x >= 0)
         return sqrt(2*pi)*x^2
       else 
         return 0
       end
    end")

# Second, we implement a MC procedure on the Julia side
julia_command("using Distributions, Random")
julia_command(
    "function integrate_mc(fcn::Function, n::Int64, seed::Int64 = 12345)
        Random.seed!(seed)
        x = rand(Normal(0, 1), n)
        fx = zeros(Float64, n)
        for r = 1:length(fx)
            fx[r] = fcn(x[r])
        end
        return mean(fx)
     end")  

# Finally, we calculate integral value (with N=100,000)
  IJ <- julia_eval("integrate_mc(fcn, 100000)")
```

The following table (Tab \@ref(tab:compare)) contains results of the integration. 

```{r compare, echo=FALSE}
library(magrittr)
library(tibble)
tribble(
  ~`Exact value`, ~`MC (R)`, ~`MC (Julia)`,
  1.25331, round(IR, 5), round(IJ, 5)
) %>% 
  knitr::kable(caption = "Integration results.", aligne = 'center')
```

As we can see, numerical results are pretty close to the actual result. Let us compare relative performance (`R`/`Julia`) vs. number of samples taken $n$ (Fig. \@ref(fig:performance)).

```{r performance, echo=FALSE, fig.width=6, fig.height=4, fig.cap="R vs. Julia performance."}
library(ggplot2)
invisible(julia_eval(paste0("integrate_mc(fcn, ", 50, ")")))
purrr::map_df(c(1000, 10000, 100000, 1000000), ~ {
  n <- as.integer(.x)
  julia_command(paste0("Random.seed!(", n, ")"))
  out <- microbenchmark::microbenchmark(
    "R" = integrate_mcR(fcnR, n), 
    "Julia" = julia_eval(paste0("integrate_mc(fcn, ", n, ")"))
  )
  tibble::tibble(
    "n" = n,
    "R" = mean(out$time[out$expr == "R"]),
    "Julia" = mean(out$time[out$expr == "Julia"]),
    "R/Julia" = `R`/`Julia`
  )
}) %>%
  ggplot(aes(x = n, y = `R/Julia`))+
  geom_point(size = 2.25)+
  geom_line(size = 0.75)+
  ylab("Time (R)/Time (Julia)")+
  scale_x_log10(breaks = c(1000, 10000, 100000, 1000000))+
  theme_bw()+
  theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12)
  )
```

As one can see, `Julia` performs much better than `R` with a sample size of MC sample increasing. 

## Writing a function based on `Julia` code

Assume that we have a `Julia` script in a file `mc-integration.jl`:

```
usign Random
using Distribution

function integrate_mc(fcn::Function, n::Int64, seed::Int64 = 12345)
  Random.seed!(seed)
  x = rand(Normal(0, 1), n)
  fx = zeros(Float64, n)
  for r = 1:length(fx)
    fx[r] = fcn(x[r])
  end
      
  return mean(fx)
end
```

Then, we can source this file using the command:

```{r, eval = FALSE}
julia_source('mc-integration.jl')
```

and create an `R` function calling `Julia` `integrate_mc()` function:

```{r}
integrate_mcJ <- function(fcn, n, seed = 12345) {
  julia_call("integrate_mc", fcn, as.integer(n), as.integer(seed))
}

integrate_mcJ(fcnR, 100000)
```



